
---
output:
  html_document:
    df_print: paged
vignette: "%\\VignetteIndexEntry{P-model usage} %\\VignetteEngine{knitr::rmarkdown}
  %\\usepackage[utf8]{inputenc}\n"
---

```{r}
# library and data loading
library(rsofun)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(cowplot)

driver <- read_rds("/data_2/FluxDataKit/v3.4/zenodo_upload/rsofun_driver_data_v3.4.rds") # currently only on workstation

plot_eval <- function(df, is_gpp = TRUE, return_gg1 = FALSE, return_gg2 = FALSE){
  
  use_col <- ifelse(is_gpp, "tomato", "royalblue")
  use_lab <- ifelse(
    is_gpp,
    expression(paste("GPP (g C m"^-2, "s"^-1, ")")), 
    expression(paste("LE (W m"^-2, ")"))
    )
  use_lab_obs <- ifelse(
    is_gpp,
    expression(paste("Observed GPP (g C m"^-2, "s"^-1, ")")), 
    expression(paste("Observed LE (W m"^-2, ")"))
    )
  use_lab_mod <- ifelse(
    is_gpp,
    expression(paste("Modeled GPP (g C m"^-2, "s"^-1, ")")), 
    expression(paste("Modeled LE (W m"^-2, ")"))
    )
  lims <- ifelse(
    is_gpp,
    c(-0.5, 10),
    c(-20, 175)
  )
   
  # Plot GPP observed and modelled quartiles by day-of-year
  gg1 <- df |> 
    tidyr::pivot_longer(c(mod, obs), names_to = "type", values_to = "var") |> 
    mutate(doy = lubridate::yday(date)) |> 
    group_by(doy, type) |> 
    summarise(
      var_q25 = quantile(var, probs = c(0.25), na.rm = TRUE),
      var_q75 = quantile(var, probs = c(0.75), na.rm = TRUE)
      ) |> 
    ggplot() +
    geom_ribbon(
      aes(
        x = doy,
        ymin = var_q25,
        ymax = var_q75,
        fill = type
        ),
      alpha = 0.75
    ) +
    scale_fill_manual(
      name = "",
      values = c(
        "mod" = use_col,
        "obs" = 'grey20'
        ),
      labels = c(
        "mod" = "Mod.",
        "obs" = "Obs."
      )) +
    theme_classic() +
    theme() +
    labs(
      x = 'Date',
      y = use_lab,
      colour = ""
    )
  
  # get fit metrics
  rsq_val <- yardstick::rsq(df, mod, obs) |> pull(.estimate)
  rmse_val <- yardstick::rmse(df, mod, obs) |> pull(.estimate)
  bias_val <- mean(df$mod - df$obs, na.rm = TRUE)
  linmod <- lm(obs ~ mod + 0, data = df)
  slope_val <- coef(linmod)
  n_val <- df |> tidyr::drop_na() |> nrow()
  
  rsq_lab <- format(rsq_val, digits = 3)
  rmse_lab <- format(rmse_val, digits = 3)
  bias_lab <- format(bias_val, digits = 3)
  slope_lab <- format(slope_val, digits = 3)
  n_lab <- n_val
  
  subtitle <- bquote(italic(R)^2 == .(rsq_lab) ~ ~
                    RMSE == .(rmse_lab) ~ ~
                    bias == .(bias_lab) ~ ~
                    slope == .(slope_lab) ~ ~
                    italic(N) == .(n_lab))
  
  # plot modelled vs. observed (actually the reverse)
  gg2 <- df |> 
    ggplot(aes(x = mod, y = obs)) +
    geom_hex(bins = 50, show.legend = FALSE) +
    theme_classic() +
    geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
    geom_hline(yintercept = 0, linetype = "dotted") +
    geom_vline(xintercept = 0, linetype = "dotted")+ 
    geom_smooth(method = "lm", formula = y ~ x + 0, color = "red", size = 0.5, se = FALSE) +
    xlim(lims[1], lims[2]) + 
    ylim(lims[1], lims[2]) +
    labs(
      x = use_lab_mod,
      y = use_lab_obs,
      subtitle = subtitle
    )
  
  if (is_gpp){
    gg2 <- gg2 +
      khroma::scale_fill_batlowW(trans = "log", reverse = TRUE)
  } else {
    gg2 <- gg2 +
      khroma::scale_fill_davos(trans = "log", reverse = TRUE)

  }
  
  if (return_gg1){
    return(gg1)
  } else if (return_gg2){
    return(gg2)
  } else {
    cow <- cowplot::plot_grid(gg1, gg2, nrow = 1)
    return(cow)
  }
}
```

## Scope

In this Vignette I'll modified the cost function used in 'calib_sofun' to calibrate the parameters using only the daily gpp and le value with a quality flag higher than 0.8. each modification of the original function are indicated with the comment "MODIFIED"

```{r}
modified_cost_likelihood_pmodel <- function(
    par,   # model parameters & error terms for each target
    obs,
    drivers,
    targets,
    par_fixed = NULL,   # non-calibrated model parameters
    parallel = FALSE,
    ncores = 2
){
  # NOTE(fabian): These different cost functions share a LOT of code in common. Consider consolidation for maintainability?
  
  # predefine variables for CRAN check compliance
  sitename <- data <- gpp_mod <- NULL
  
  if (!("use_phydro" %in% colnames(drivers$params_siml[[1]]))){
    warning("Parameter use_phydro not set. Assuming FALSE")
    using_phydro = FALSE
  } else {
    using_phydro = drivers$params_siml[[1]]$use_phydro
  }
  
  ## define required parameter set based on model parameters
  if (!using_phydro){
    required_param_names <- rsofun:::required_param_names$p_model
  } else {
    required_param_names <- rsofun:::required_param_names$phydro_model
  }
  
  ## split calibrated parameters into model and error parameters
  par_calibrated_model      <- par[!startsWith(names(par), "err_")] # consider only model parameters for the check
  # par_calibrated_errormodel <- par[   names(par) %in% c("err_gpp", "err_vcmax25") ]
  # par_fixed
  
  ## check parameters
  if (!identical(sort(c(names(par_calibrated_model), names(par_fixed))), required_param_names)){
    stop(sprintf(paste0("Error: Input calibratable and fixed parameters do not ",
                        "match required model parameters:",
                        "\n         par:       c(%s)",
                        "\n         par_fixed: c(%s)",
                        "\n         required:  c(%s)"),
                 paste0(sort(names(par_calibrated_model)), collapse = ", "),
                 paste0(sort(names(par_fixed)), collapse = ", "),
                 paste0(sort(required_param_names), collapse = ", ")))
  }
  
  # Combine fixed and estimated params to result in all the params required to run the model
  # This basically uses all params except those of the error model of the observations
  params_modl <- c(par, par_fixed)[required_param_names]
  
  ## run the model
  df <- runread_pmodel_f(
    drivers,
    par = params_modl,
    makecheck = TRUE,
    parallel = parallel,
    ncores = ncores
  )
  
  ## clean model output and unnest
  df <- df |>
    dplyr::rowwise() |>
    dplyr::reframe(
      cbind(sitename, data[, c('date', unique(c('gpp', targets)))]) |>
        stats::setNames(c('sitename', 'date', paste0(unique(c('gpp', targets)), '_mod')))
    ) # gpp is used to get average trait prediction
  
  # separate validation data into fluxes and traits, site by site
  is_flux <- apply(obs, 1, function(x){ 'date' %in% colnames(x$data)})
  
  if(sum(is_flux) > 0){
    flux_sites <- obs$sitename[is_flux]
    
    # Unnest flux observations for our targets
    obs_flux <- obs[is_flux, ] |>
      dplyr::select(sitename, data) |>
      tidyr::unnest(data) |>
      dplyr::select(any_of(c('sitename', 'date', targets,paste0(targets,"_qc"))))
    
    if(ncol(obs_flux) < 3){
      warning("Dated observations (fluxes) are missing for the chosen targets.")
      df_flux <- data.frame()
    }else{
      # Join P-model output and flux observations
      df_flux <- df |>
        dplyr::filter(sitename %in% flux_sites) |>
        dplyr::left_join(
          obs_flux, 
          by = c('sitename', 'date'))    # observations with missing date are ignored
    }
  }else{
    df_flux <- data.frame()
  }
  
  if(sum(!is_flux) > 0){
    trait_sites <- obs$sitename[!is_flux]
    
    # Unnest trait observations for our targets
    obs_trait <- obs[!is_flux, ] |>
      dplyr::select(sitename, data) |>
      tidyr::unnest(data) |>
      dplyr::select(any_of(c('sitename', targets,paste0(targets,"_qc"))))
    
    if(ncol(obs_trait) < 2){
      warning("Non-dated observations (traits) are missing for the chosen targets.")
      df_trait <- data.frame()
    }else{
      # Join output and trait observations
      df_trait <- df |>
        dplyr::filter(sitename %in% trait_sites) |>
        dplyr::group_by(sitename) |>
        # get growing season average traits
        dplyr::summarise(across(ends_with("_mod") & !starts_with('gpp'),
                                ~ sum(.x * gpp_mod/sum(gpp_mod)),
                                .names = "{.col}")) |>
        dplyr::left_join(
          obs_trait,
          by = c('sitename')        # compare yearly averages rather than daily obs
        )
    }
  }else{
    df_trait <- data.frame()
  }
  
  # loop over targets to compute log-likelihood ll
  ll_df <- data.frame(target = targets, 
                      ll     = NaN)
  for (target in targets){
    # check (needed?):
    if(target %in% colnames(df_flux) & target %in% colnames(df_trait)) {stop(
      sprintf("Target '%s' cannot be simultatneously in df_flux and df_trait.", target))
    }
    
    # get observations and predicted target values, without NA 
    df_target <- if(target %in% colnames(df_flux)){
      df_flux[, c(paste0(target, '_mod'), target ,paste0(target, '_qc'))] |> tidyr::drop_na()
    }else{
      df_trait[, c(paste0(target, '_mod'), target, paste0(target, '_qc'))] |> tidyr::drop_na()
    }
    
    # le in the output is on d^-1 while in input is in s^-1
    if(target == "le"){
      df_target$le_mod <- df_target$le_mod / (24*60*60)
    }
    
    df_target <- df_target |> filter(if_all(ends_with("_qc"), ~ . > 0.8))
    
    # calculate normal log-likelihood
    ll_df[ll_df$target == target, 'll'] <- 
      sum(stats::dnorm(
        x    = df_target[[paste0(target, '_mod')]], # model
        mean = df_target[[target]],                 # obs
        sd   = par[[paste0('err_', target)]],       # error model
        log  = TRUE))
  }
  ll <- sum(ll_df$ll)
  
  # trap boundary conditions
  if(is.nan(ll) | is.na(ll) | ll == 0){ll <- -Inf}
  
  return(ll)
}

```

I'll use the site FR-Pue. I'll run the simulation using the Penmann-Montheith equation using non-calibrated and calibrated parameter

```{r}

# select the site with the highest quality data

single_site <- driver[driver$sitename == "FR-Pue",]

single_site <- single_site |> unnest(params_siml) |>
  mutate(use_gs = TRUE,
         use_phydro = FALSE,
         use_pml= TRUE) |>
  group_by(sitename) |>
  nest(params_siml = c(spinup, spinupyears, recycle, outdt, ltre,  ltne,  ltrd,  ltnd,  lgr3,  lgn3,  lgr4 ,
                       use_gs, use_phydro, use_pml))


single_site$site_info[[1]]$whc <- 255
```


## Calibration 

approx. 30 min

```{r}
# LE and GPP calibration

evaluation <- single_site |>
  unnest(forcing) |>
  dplyr::select(sitename,date,gpp,gpp_qc,le,le_qc) |>
  group_by(sitename) |>
  nest(data = c(date,gpp,gpp_qc,le,le_qc))

params_fix <- list(
  rd_to_vcmax        = 0.014,
  tau_acclim         = 30)

settings <- list(
  method = "BayesianTools",
  par = list(
    kphio = list(lower=0.03, upper=0.2, init = 0.05),
    kphio_par_a = list(lower = -0.0004, upper = 0.000001, init = -0.0025),
    kphio_par_b = list(lower = 10, upper = 30, init = 20),
    soilm_thetastar = list(lower  = 0, upper = 240, init = 144),
    beta_unitcostratio = list(lower = 5, upper = 200, init= 50),
    kc_jmax            =  list(lower = 0.2, upper = 0.8, init = 0.41),
    gw_calib = list(lower = 0.5, upper = 10, init = 2),
    err_gpp = list(lower = 0.01, upper = 4, init = 2),
    err_le = list(lower = 0.01, upper = 170, init = 38)
  ),
  metric = modified_cost_likelihood_pmodel,
  control = list(
    sampler = "DEzs",
    settings = list(
      burnin = 5000,
      iterations = 20000,
      nrChains = 3,        # number of independent chains
      startValue = 3    # kept artificially low,
    )
  )
)

# Run the calibration for GPP and le 
calib_output <- rsofun::calib_sofun(
  drivers = single_site,
  obs = evaluation,
  settings = settings,
  # extra arguments for the cost function
  par_fixed = params_fix,
  targets = c("gpp",  "le")
)

```

## Calibration results

```{r, fig.height=10, fig.width= 16}
plot(calib_output$mod)

```



## Run with calibrated parameter


I run the calibration separately and insert manually the results

```{r, fig.width=8, fig.height=4}
params_modl <- list(
  kphio              =  calib_output$par[["kphio"]],   
  kphio_par_a        =  calib_output$par[["kphio_par_a"]],       
  kphio_par_b        =  calib_output$par[["kphio_par_b"]],       
  rd_to_vcmax        =  0.014,  
  soilm_thetastar    =  calib_output$par[["soilm_thetastar"]],       
  beta_unitcostratio =  calib_output$par[["beta_unitcostratio"]],       
  tau_acclim         =  30,
  kc_jmax            =  calib_output$par[["kc_jmax"]],
  gw_calib = calib_output$par[["gw_calib"]]#calib_output$par[["kc_jmax"]]
)



output <- rsofun::runread_pmodel_f(
  single_site,
  par = params_modl
)


df_plot <- output |>
  unnest(data) |>
  select(date, mod = gpp) |> 
  left_join(
    single_site |>
      unnest(forcing) |> 
      select(date, obs = gpp),
    by = join_by(date)
  )

plot_eval(df_plot)
```

The GPP prediction are very good, however the LE prediction are worse using the new parameters



### LE results

```{r, fig.width=8, fig.height=4}

df_plot <- output |>
  unnest(data) |>
  mutate(le =  le / (24 * 60 * 60)) |> 
  select(date, mod = le) |> 
  left_join(
    p_model_validation |>
      unnest(data) |> 
      select(date, obs = le),
    by = join_by(date)
  )

plot_eval(df_plot, is_gpp = FALSE)
```
